Foundations of ecological and evolutionary change

### 2025-06-08 foundations_of_ecoevo.tex
Manuscript updates

The interdependence of ecological and evolutionary processes has long
been recognised (e.g., \citeproc{ref-Darwin1859}{Darwin 1859};
\citeproc{ref-Fisher1958}{Fisher 1958};
\citeproc{ref-Pelletier2009}{Pelletier et al. 2009}), but the rise of
eco-evolutionary models, which incorporate both, is relatively recent
following a widespread recognition that ecology and evolution can happen
on similar timescales (\citeproc{ref-Govaert2019}{Govaert et al. 2019};
\citeproc{ref-Yamamichi2023}{Yamamichi et al. 2023}). Currently, a
universally recognised formal definition of eco-evolutionary change is
lacking, with some theoreticians broadly interpreting ``eco-evolutionary
dynamics'' to allow for a separation of ecological and evolutionary
timescales (\citeproc{ref-Lion2023}{Lion et al. 2023}) and others
advocating for a more narrow interpretation in which no such separation
is permitted (\citeproc{ref-Bassar2021}{Bassar et al. 2021}). Bassar et
al. (\citeproc{ref-Bassar2021}{2021}) identify two types of
eco-evolutionary models that follow from these interpretations. The
first type uses separate equations to model population change versus
evolutionary change, thereby allowing for any number of ecological,
evolutionary, or environmental feedbacks (e.g.,
\citeproc{ref-Lion2018}{Lion 2018}; \citeproc{ref-Patel2018}{Patel et
al. 2018}; \citeproc{ref-Lion2023}{Lion et al. 2023}). The second type
models population demographics as functions of quantitative traits, with
ecological and evolutionary change following from demographic processes
and trait distributions operating on the same timescale (e.g.,
\citeproc{ref-Barfield2011}{Barfield et al. 2011};
\citeproc{ref-Simmonds2020}{Simmonds et al. 2020};
\citeproc{ref-Jaggi2024}{Jaggi et al. 2024}). Both model types can be
very general, but like all predictive models, they rely on simplifying
assumptions for tractability (\citeproc{ref-Levins1966}{Levins 1966};
\citeproc{ref-Luque2016}{Luque 2017}). These simplifying assumptions are
often grounded in the Price equation to demonstrate accuracy and logical
consistency when modelling evolutionary change (e.g.,
\citeproc{ref-Coulson2008}{Coulson and Tuljapurkar 2008};
\citeproc{ref-Barfield2011}{Barfield et al. 2011};
\citeproc{ref-Rees2016}{Rees and Ellner 2016};
\citeproc{ref-Lion2018}{Lion 2018}). For example, Barfield et al.
(\citeproc{ref-Barfield2011}{2011}) link their model back to Price
(\citeproc{ref-Price1970}{1970}), which they consider to be a
``universal law of evolution'', to place their conclusions concerning
stage structured evolution in the broader context of evolutionary
theory. The role of fundamental equations is therefore important for
unifying theory (\citeproc{ref-Luque2016}{Luque 2017}), and we believe
that a fundamental equation of eco-evolutionary change is needed.

@@

\begin{quote}
\textbf{Box 1:} The Price equation is an abstract formula to represent
evolutionary change. Formulated originally in the early 1970s by George
Price (\citeproc{ref-Price1970}{Price 1970},
\citeproc{ref-Price1972}{1972}), it postulates some basic properties
that all evolutionary systems must satisfy: change over time, ancestor
and descendant relations, and a character or phenotype
(\citeproc{ref-Rice2004}{Rice 2004}). Using simple algebraic language,
the Price equation represents evolutionary change with the predominant
notation,
\[\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).\]
In the above equation, \(\Delta\bar{z}\) is the change in the average
character value \(z\) over a time step of arbitrary length, \(w\) is an
individual's absolute fitness, and \(\bar{w}\) average population
fitness. On the right-hand side of the equation, the first term is the
covariance between a character value \(z\) and fitness \(w\), which
reflects \(\bar{z}\) change attributable to differential survival and
reproduction. The second term is the expected value of \(\Delta z\),
which reflects the extent to which offspring deviate from parents in
\(z\) (\citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Okasha2006}{Okasha 2006}; \citeproc{ref-Frank2012}{Frank
2012}). A more specific version of the covariance term was already known
within the quantitative and population genetics tradition
(\citeproc{ref-Robertson1966}{Robertson 1966}), usually representing the
action of natural selection. The Price equation adds an expectation term
and abstracts away from any specific mechanisms of replication or
reproduction, or mechanisms of inheritance. Its definitional nature and
lack of substantive biological assumptions has been portrayed both as a
strength (\citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Frank2012}{Frank 2012}; \citeproc{ref-Luque2016}{Luque
2017}; \citeproc{ref-Baravalle2022}{Baravalle and Luque 2022}), and its
greatest weakness. The abstract nature of the Price equation places it
at the top of the hierarchy of fundamental theorems of evolution from
which the rest (Robertson's theorem, Fisher's fundamental theorem,
breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be
easily derived (\citeproc{ref-Lehtonen2016}{Lehtonen 2016},
\citeproc{ref-Lehtonen2018}{2018}; \citeproc{ref-Queller2017}{Queller
2017}). This abstractness is also key to developing a more general view
of evolution (\citeproc{ref-Rice2020}{Rice 2020};
\citeproc{ref-Luque2021}{Luque and Baravalle 2021};
\citeproc{ref-Edelaar2023}{Edelaar et al. 2023}). In contrast, some
researchers consider the Price equation just a triviality (even
tautological), and useless without further modelling assumptions
(\citeproc{ref-VanVeelen2005}{van Veelen 2005};
\citeproc{ref-VanVeelen2012}{van Veelen et al. 2012}). The debate
remains open (\citeproc{ref-VanVeelen2020}{van Veelen 2020};
\citeproc{ref-Baravalle2024}{Baravalle et al. 2024}).
\end{quote}

@@

The foundation of eco-evolutionary change is an interaction between the
demographic processes of birth and death
(\(\beta_{i} - \delta_{i} + 1\)) and some characteristic of an
individual (\(z_{i} + \Delta z_{i}\)). The value of \(\Omega\) is a
summation of individual characteristics, which takes the same units as
\(z\). From eqn 1, we can derive the most fundamental equations of
population ecology (Box 2) and evolutionary biology (Box 1) through an
appropriate interpretation of \(z\). Under many interpretations of
\(z\), we can also interpret \(\Omega\) as a metric of ecosystem
function.

@@

We can now interpret \(\Omega\) as the population size at \(t+1\),
\(N_{t+1}\) and note that \(N\beta\), \(N\delta\), and \(N\) are the
total births, total deaths, and size of the population at \(t\),
respectively,

@@

If we assume that individuals are identical, then we can drop the
subscript \(i\) such that \(\beta_{i} = \beta\) and
\(\delta_{i} = \delta\). Summing from 1 to \(N\), we can rewrite the
above,

@@

In some cases, we might also consider \(\Omega\) to be the total
contribution of a population to ecosystem function. For example, \(z\)
might be interpreted as trait that underlies a function of interest such
as carbon capture or photosynthetic rate. When that the relevant
characteristic (\(z_{i}\)) is additive across individuals (e.g., the
total amount of carbon capture in a population is the sum of carbon
captured across individuals), \(\Omega\) could be an appropriate metric
of ecosystem function. This will not always be the case. The sum of some
measurable traits such as leaf pH
(\citeproc{ref-Cornelissen2011}{Cornelissen et al. 2011}) will not be
ecologically meaningful when summed across individuals. And in other
cases, the relationship between \(\Omega\) and some ecosystem function
of interest is likely to be subadditive. For example, we could interpret
\(z_{i}\) as the frequency at which a pollinator \(i\) visits plants
(which might change from \(t\) to \(t+1\) for any number of reasons,
including ontogenic or environmental, as reflected in \(\Delta z_{i}\)).
While \(\Omega\) would then be the total frequency of pollinator visits
in a population, pollinator saturation effects would likely have a
logarithmic relationship between \(\Omega\) and total seed set (e.g.,
\citeproc{ref-Mu2018}{Mu et al. 2018}). While we have argued that
ecological and evolutionary change can both be derived from the same
fundamental relationship between birth, death, and individual
characteristics, the same is not true for ecosystem function.
Nevertheless, Box 3 provides a toy example of four plants with different
fitnesses and carbon captures. We use this toy example to demonstrate a
concrete example of our proposed framework.

@@

A classical sign of scientific progress is the ability to connect
disparate theories and models to show how empirical and theoretical
models are logical (mathematical) consequences of more fundamental ones
(\citeproc{ref-Nagel1961}{Nagel 1961}). Conceptual unification therefore
has a critical role in advancing scientific theory
(\citeproc{ref-Morrison2000}{Morrison 2000}). Showing how apparently
disparate phenomena follow from the same shared principles reveals what
is most fundamental about the natural world and provides a foundation
for further scientific investigation and the construction of more
coherent and predictive models. Our eqn 1 provides a foundation for the
conceptual unification of the fundamental equations of population
ecology and evolution.

Rather than making simplifying assumptions, as is the approach for most
ecological and evolutionary models, we focus on fundamental axioms that
are universal to closed biological systems: discrete individuals, birth,
death, and change over time. We define an abstract sum (\(\Omega\)), to
which all individuals within the population contribute. From the simple
assumptions of population identity (\(z_{i} = 1\)) and invariability
(\(\Delta z_{i} = 0\)), we recover the most general equation of
population ecology (Box 2) and principle of exponential growth
(\(N_{t+1} = N_{t}\lambda\)). By defining individual fitness (\(w_{i}\))
and applying the total conservation of probability to individual
frequencies (\citeproc{ref-Frank2015}{Frank 2015},
\citeproc{ref-Frank2016}{2016}), we recover the most fundamental
equation of evolution (Box 1). We therefore deliver a general framework
to unify fundamental equations of ecological and evolutionary change.

The Price equation provides a complete and exact description of
evolution in any closed evolving system (\citeproc{ref-Price1970}{Price
1970}; \citeproc{ref-Frank2012}{Frank 2012}). It is derived by
rearranging the mathematical notation defining changes in the
frequencies and characteristics of any type of entity
(\citeproc{ref-Price1970}{Price 1970};
\citeproc{ref-Gardner2008}{Gardner 2008}; \citeproc{ref-Luque2016}{Luque
2017}) (e.g., individuals, alleles). This derivation partitions total
characteristic change into different components, making it possible to
isolate evolutionary mechanisms (e.g., selection) and levels of
biological organisation (e.g., group, individual)
(\citeproc{ref-Frank1995}{Frank 1995}, \citeproc{ref-Frank2012}{2012};
\citeproc{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
\citeproc{ref-Luque2016}{Luque 2017}; \citeproc{ref-Okasha2020}{Okasha
and Otsuka 2020}). Because of its abstract nature and lack of any
system-specific assumptions, the Price equation makes no predictions
about what will happen in any particular system
(\citeproc{ref-Gardner2020}{Gardner 2020}). Its role is not to predict,
but to formally and completely define and separate components of
evolutionary change. The same is true of the general equation for
population change (eqns 2 and 3), at least as we have used it here where
it serves to define what population growth means in ecology. This
equation formally and completely describes population change in terms of
births and deaths. In eqn 1, we therefore have a fundamental equation
from which we can derive complete ecological and evolutionary change in
any closed biological population. We anticipate that this will be useful
for eco-evolutionary theory in the same way that the Price equation is
useful for evolutionary theory: facilitating specific model development
and identifying new conceptual insights, unresolved errors, and sources
of model disagreements.

Our unification demonstrates the equivalence between the finite rate of
increase \(\lambda\) (Box 2) (\citeproc{ref-Gotelli2001}{Gotelli 2001})
and population mean evolutionary fitness \(\bar{w}\) (Box 1). The
population growth equation \(N_{t+1} = N_{t}\lambda\) could therefore be
rewritten as \(N_{t+1} = N_{t}\bar{w}\). We do not claim to be the first
to notice this equivalence between population growth and mean population
fitness. Indeed, the relationship between population growth rate and
evolutionary fitness has been proposed and applied many times before
(\citeproc{ref-Fisher1930}{Fisher 1930};
\citeproc{ref-Charlesworth1980}{Charlesworth 1980};
\citeproc{ref-Lande1982}{Lande 1982}; \citeproc{ref-Roff2008}{Roff
2008}; \citeproc{ref-Lion2018}{Lion 2018}). For example, Lande
(\citeproc{ref-Lande1982}{1982}) explicitly concludes mean absolute
fitness per unit time is \(\bar{w} = e^{r}\), where \(r = ln(\lambda)\).
We show this from first principles and clarify the relationship between
fitness and population growth. Over an arbitrary length of time, fitness
is properly defined as \(w_{i} = \beta_{i} - \delta_{i} + 1\). Over an
individual's lifetime (which, by definition, includes death), fitness is
therefore \(\beta_{i}\). Interestingly, the rate of change in ecology
and evolution are reflected in the first and second statistical moments
of fitness, respectively. Population growth rate reflects mean fitness
\(\bar{w}\), while the rate of evolutionary change reflects the variance
in fitness \(\mathrm{Var}(w)\) (i.e., Fisher's fundamental theorem)
(\citeproc{ref-Frank1997}{Frank 1997}; \citeproc{ref-Rice2004}{Rice
2004}; \citeproc{ref-Queller2017}{Queller 2017}).

Our unification may also help explain, at least partially, some of the
success of classical population genetic models. For decades, population
genetics (and to some extent quantitative genetics) has been accused of
being a reductionistic view of evolution, reducing everything to changes
in allele frequencies and abstracting away from individuals and their
environments (the ecological interactions)
(\citeproc{ref-MacColl2011}{MacColl 2011}). This has been a line of
argumentation by some defenders of the so-called Extended Synthesis
(\citeproc{ref-Pigliucci2009}{Pigliucci 2009}), especially in relation
to niche construction (\citeproc{ref-Odling-smee2003}{Odling-Smee et al.
2003}). Famously, Mayr (\citeproc{ref-Mayr1960}{1959}) characterized
population genetics as a simple input and output of genes, analogous to
``the adding of certain beans to a beanbag and the withdrawing of
others'' (also called ``beanbag genetics'').

Historical critics of population genetics could not articulate a clear
explanation for why it works so well despite all of its idealisations
and simplifications. From the Price equation, we are able to recover
classical population and quantitative genetic models
(\citeproc{ref-Queller2017}{Queller 2017}) and develop new ones
(\citeproc{ref-Rice2004}{Rice 2004}, \citeproc{ref-Rice2020}{2020};
\citeproc{ref-Luque2016}{Luque 2017}; \citeproc{ref-Lion2018}{Lion
2018}). Our eqn 1 contains ecology at its core, and we show how the
Price equation logically follows from it after accounting for absolute
population growth (eqn 8). We therefore conclude that population and
quantitative genetic equations contain ecology (no matter how hidden),
and the ecological nature of evolution is implicit in population and
quantitative genetic models.

@@

Our fundamental equation is complete and exact. It therefore implicitly
includes any effects of density dependence on population growth (see Box
2), or any social effects on evolutionary change (see Box 1). We can
make both of these effects explicit with the same partition in the
summation on the right-hand side of eqn 1. To do this, we multiply the
expression
\(\left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)\)
by
\(\left(1 - \sum_{j=1}^{N} a_{ij}\left(z_{j} + \Delta z_{j}\right)\right)\),
where \(j\) indexes the same individuals as \(i\), and \(a_{ij}\)
defines the effect \(j\) has on the contribution of \(i\) to \(\Omega\).
In evolutionary terms, \(a_{ij}\) defines how \(j\) modulates the
fitness of \(i\) (\(w_{i}\)). In ecological terms, \(a_{ij}\) defines
how \(j\) modulates the contribution of \(i\) to total population
growth. If we define \(\alpha\) as the summed effect of all individuals
on a focal \(i\) (i.e., \(\alpha = \sum_{j=1}^{N} a_{ij}\)), then we can
partition ecological change into density-independent and
density-dependent effects and derive a discrete time logistic growth
equation. We save these derivations for a future investigations focused
on evolutionary and community ecology.

We have shown that we can derive the fundamental equations of population
ecology and biological evolution from a single unifying equation.
Lastly, we propose our eqn 1 as a potential starting point for defining
ecosystem function and further conceptual unification between ecology,
evolution, and ecosystem function. The Price equation has previously
been used to investigate ecosystem function
(\citeproc{ref-Loreau2001}{Loreau and Hector 2001};
\citeproc{ref-Fox2006}{Fox 2006}), but not with any attempt towards
conceptual unification with evolutionary biology. For example, Fox
(\citeproc{ref-Fox2006}{2006}) applied the abstract properties of the
Price equation to partition total change in ecosystem function into
separate components attributable to species richness, species
composition, and context dependent effects. This approach provides a
framework for comparing the effects of biodiversity on ecosystem
function in empirical systems (\citeproc{ref-Fox2006}{Fox 2006};
\citeproc{ref-Winfree2015}{Winfree et al. 2015};
\citeproc{ref-Mateo-Tomas2017}{Mateo-Tomás et al. 2017}). Instead, our
eqn 1 defines \(\Omega\) as total ecosystem function contributed by a
focal population (e.g., biomass production, decomposition,
photosynthetic rate). It is therefore possible to investigate
ecological, evolutionary, and ecosystem function change from the same
shared framework.

@@

\bibitem[\citeproctext]{ref-Baravalle2024}
Baravalle, L., A. Roffé, V. J. Luque, and S. Ginnobili. 2024.
\href{https://doi.org/10.1007/s13752-024-00482-4}{The value of price}.
Biological Theory.

@@

\bibitem[\citeproctext]{ref-Cornelissen2011}
Cornelissen, J. H. C., F. Sibma, R. S. P. V. Logtestijn, R. A. Broekman,
and K. Thompson. 2011.
\href{https://doi.org/10.1111/j.1365-2435.2010.01765.x}{Leaf pH as a
plant trait: Species-driven rather than soil-driven variation}.
Functional Ecology 25:449--455.

@@

\bibitem[\citeproctext]{ref-Mu2018}
Mu, J., Q. Wu, Y. Yang, M. Huang, and C. M. Grozinger. 2018.
\href{https://doi.org/10.1111/oik.04711}{Plant reproductive strategies
vary under low and high pollinator densities}. Oikos 127:1081--1094.

>>>>

The interdependence of ecological and evolutionary processes has long
been recognised (e.g., \citeproc{ref-Darwin1859}{Darwin 1859};
\citeproc{ref-Fisher1958}{Fisher 1958};
\citeproc{ref-Pelletier2009}{Pelletier et al. 2009}), but the rise of
eco-evolutionary models, which incorporate both, is relatively recent
following a widespread recognition that ecology and evolution can happen
on similar timescales (\citeproc{ref-Govaert2019}{Govaert et al. 2019};
\citeproc{ref-Yamamichi2023}{Yamamichi et al. 2023}). Currently, a
universally recognised formal definition of eco-evolutionary change is
lacking, with some theoreticians broadly interpreting ``eco-evolutionary
dynamics'' to allow for a separation of ecological and evolutionary
timescales (\citeproc{ref-Lion2023}{Lion et al. 2023}) and others
advocating for a more narrow interpretation in which no such separation
is permitted (\citeproc{ref-Bassar2021}{Bassar et al. 2021}). Bassar et
al. (\citeproc{ref-Bassar2021}{2021}) identify two types of
eco-evolutionary models that follow from these interpretations. The
first type uses separate equations to model population change versus
evolutionary change, thereby allowing for any number of ecological,
evolutionary, or environmental feedbacks (e.g.,
\citeproc{ref-Lion2018}{Lion 2018}; \citeproc{ref-Patel2018}{Patel et
al. 2018}; \citeproc{ref-Lion2023}{Lion et al. 2023}). The second type
models population demographics as functions of quantitative traits, with
ecological and evolutionary change following from demographic processes
and trait distributions operating on the same timescale (e.g.,
\citeproc{ref-Barfield2011}{Barfield et al. 2011};
\citeproc{ref-Simmonds2020}{Simmonds et al. 2020};
\citeproc{ref-Jaggi2024}{Jaggi et al. 2024}). Both model types can be
very general, but like all predictive models, they rely on simplifying
assumptions for tractability (\citeproc{ref-Levins1966}{Levins 1966};
\citeproc{ref-Luque2016}{Luque 2017}). These simplifying assumptions are
often grounded in the Price equation to demonstrate accuracy and logical
consistency when modelling evolutionary change (e.g.,
\citeproc{ref-Coulson2008}{Coulson and Tuljapurkar 2008};
\citeproc{ref-Barfield2011}{Barfield et al. 2011};
\citeproc{ref-Rees2016}{Rees and Ellner 2016};
\citeproc{ref-Lion2018}{Lion 2018}). For example, Barfield et al.
(\citeproc{ref-Barfield2011}{2011}) link their model back to Price
(\citeproc{ref-Price1970}{1970}), which they consider to be a
``universal law of evolution'', to place their conclusions concerning
stage structured evolution in the broader context of evolutionary
theory. The role of fundamental equations is therefore important for
unifying theory (\citeproc{ref-Luque2021}{Luque and Baravalle 2021}),
and we believe that a fundamental equation of eco-evolutionary change is
needed.

@@

\begin{quote}
\textbf{Box 1:} The Price equation is an abstract formula to represent
evolutionary change. Formulated originally in the early 1970s by George
Price (\citeproc{ref-Price1970}{Price 1970},
\citeproc{ref-Price1972}{1972}), it postulates some basic properties
that all evolutionary systems must satisfy: change over time, ancestor
and descendant relations, and a character or phenotype
(\citeproc{ref-Rice2004}{Rice 2004}). Using simple algebraic language,
the Price equation represents evolutionary change with the predominant
notation,
\[\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).\]
In the above equation, \(\Delta\bar{z}\) is the change in the average
character value \(z\) over a time step of arbitrary length, \(w\) is an
individual's fitness, and \(\bar{w}\) average population fitness. On the
right-hand side of the equation, the first term is the covariance
between a character value \(z\) and fitness \(w\), which reflects
\(\bar{z}\) change attributable to differential survival and
reproduction. The second term is the expected value of \(\Delta z\),
which reflects the extent to which offspring deviate from parents in
\(z\) (\citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Okasha2006}{Okasha 2006}; \citeproc{ref-Frank2012}{Frank
2012}). A more specific version of the covariance term was already known
within the quantitative and population genetics tradition
(\citeproc{ref-Robertson1966}{Robertson 1966}), usually representing the
action of natural selection. The Price equation adds an expectation term
and abstracts away from any specific mechanisms of replication or
reproduction, or mechanisms of inheritance. Its definitional nature and
lack of substantive biological assumptions has been portrayed both as a
strength (\citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Frank2012}{Frank 2012}; \citeproc{ref-Luque2016}{Luque
2017}; \citeproc{ref-Baravalle2022}{Baravalle and Luque 2022}), and its
greatest weakness. The abstract nature of the Price equation places it
at the top of the hierarchy of fundamental theorems of evolution from
which the rest (Robertson's theorem, Fisher's fundamental theorem,
breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be
easily derived (\citeproc{ref-Lehtonen2016}{Lehtonen 2016},
\citeproc{ref-Lehtonen2018}{2018}; \citeproc{ref-Queller2017}{Queller
2017}). This abstractness is also key to developing a more general view
of evolution (\citeproc{ref-Rice2020}{Rice 2020};
\citeproc{ref-Luque2021}{Luque and Baravalle 2021};
\citeproc{ref-Edelaar2023}{Edelaar et al. 2023}). In contrast, some
researchers consider the Price equation just a triviality (even
tautological), and useless without further modelling assumptions
(\citeproc{ref-VanVeelen2005}{van Veelen 2005};
\citeproc{ref-VanVeelen2012}{van Veelen et al. 2012}). The debate
remains open (\citeproc{ref-VanVeelen2020}{van Veelen 2020};
\citeproc{ref-Baravalle2024}{Baravalle et al. 2025}).
\end{quote}

@@

The foundation of eco-evolutionary change is an interaction between the
demographic processes of birth and death
(\(\beta_{i} - \delta_{i} + 1\)) and some characteristic of an
individual (\(z_{i} + \Delta z_{i}\)). The value of \(\Omega\) is a
summation of individual characteristics, which takes the same units as
\(z\). From eqn 1, we can derive the most fundamental equations of
population ecology (Box 2) and evolutionary biology (Box 1) through an
appropriate interpretation of \(z\). Under some limited interpretations
of \(z\), we can also interpret \(\Omega\) as a metric of ecosystem
function.

@@

We can now interpret \(\Omega\) as the population size at \(t+1\),
\(N_{t+1}\),

@@

If we assume that individuals are identical, then we can drop the
subscript \(i\) such that \(\beta_{i} = \beta\) and
\(\delta_{i} = \delta\) and note that \(N\beta\), \(N\delta\), and \(N\)
are the total births, total deaths, and size of the population at \(t\),
respectively,

Summing from 1 to \(N\), we can rewrite the above,

@@

In some cases, \(\Omega\) could also be intepreted as the total
contribution of a population to ecosystem function. This will be
restricted to cases in which \(z\) is an individual characteristic
defining an absolute quantity measured at the whole organism level, such
as biomass, seed production, carbon capture, flower visits, or nutrient
consumption. In these cases, the sum across individuals gives a
meaningful total quantity for the population. When \(z\) is instead
defined by relative organism-level measurements such wing loading,
nutrient ratio, or diet composition, or when \(z\) is measured at a
level of biological organisation below the organism (e.g., average cell
volume or leaf surface area), \(\Omega\) does not have a clear
population-level interpretation.

Box 3 provides a toy example of three plants with different fitnesses
and carbon captures. We use this toy example to demonstrate a concrete
example of our proposed framework.

\begin{quote}
\textbf{Box 3:} As a toy example of our framework, consider a population
of \(N_{t} = 3\) annual plants in which individual fruit mass (kg) is
measured, and change is observed over a year. For all plants,
\(\delta_{i} = 1\), and let plant fecundities be \(\beta_{1} = 1\),
\(\beta_{2} = 1\), and \(\beta_{3} = 2\). Applying eqn 1 to population
change such that \(\Omega = N_{t+1}\), \(z_{i} = 1\) and
\(\Delta z_{i} = 0\),
\(N_{t+1} = (1 - 1 + 1)(1 + 0) + (1 - 1 + 1)(1 + 0) + (2 - 1 + 1)(1 + 0) = 4\)
(note \(\bar{w} = 4/3\), so \(N_{t}\bar{w} = 4\)). Focusing next on the
characteristic of total fruit mass, let \(z_{1} = 0.8\),
\(z_{2} = 1.0\), and \(z_{3} = 1.5\). Also let \(\Delta z_{i} = 0.1\)
for all plants to reflect a change in soil environment from \(t\) to
\(t + 1\). In this case, total fruit yield at \(t+1\) is
\(\Omega = (1 - 1 + 1)(0.8 + 0.1) + (1 - 1 + 1)(1.0 + 0.1) + (2 - 1 + 1)(1.5 + 0.1) = 5.2\)
kg. At \(t\), mean fruit yield per plant was 1.1 kg, but at \(t+1\),
mean fruit yield per plant is \(1.3\) kg. Note that
\(\mathrm{Cov}(w, z) = 2/15\) and \(\mathrm{E}(w \Delta z) = 2/15\), so
applying the Price equation, \(\bar{w}\Delta z = 2/15 + 2/15 = 4/15\).
Since \(\bar{w} = 4/3\), multiplying both sides of the equation by
\(3/4\) returns \(\Delta z = 0.2\), which is the mean difference in
fruit yield between \(t+1\) and \(t\). The framework expressed in eqn 1
thereby links population change, evolutionary change, and ecosystem
function.
\end{quote}

@@

A classical sign of scientific progress is the ability to connect
disparate theories and models to show how empirical and theoretical
models are logical (mathematical) consequences of more fundamental ones
(\citeproc{ref-Nagel1961}{Nagel 1961};
\citeproc{ref-Morrison2000}{Morrison 2000}). Rather than making
simplifying assumptions, as is the approach for most ecological and
evolutionary models, we focus on fundamental axioms that are universal
to closed biological systems: discrete individuals, birth, death, and
change over time. We define an abstract sum (\(\Omega\)), to which all
individuals within the population contribute. From the simple
assumptions of population identity (\(z_{i} = 1\)) and invariability
(\(\Delta z_{i} = 0\)), we recover the most general equation of
population ecology (Box 2) and principle of exponential growth
(\(N_{t+1} = N_{t}\lambda\)). By defining individual fitness (\(w_{i}\))
and applying the total conservation of probability to individual
frequencies (\citeproc{ref-Frank2015}{Frank 2015},
\citeproc{ref-Frank2016}{2016}), we recover the most fundamental
equation of evolution (Box 1). Our eqn 1 thereby provides a foundation
for defining eco-evolutionary change in any population.

The Price equation provides a complete and exact description of
evolution in any closed evolving system (\citeproc{ref-Price1970}{Price
1970}; \citeproc{ref-Frank2012}{Frank 2012}). It is derived by
rearranging the mathematical notation defining changes in the
frequencies and characteristics of any type of entity
(\citeproc{ref-Price1970}{Price 1970};
\citeproc{ref-Gardner2008}{Gardner 2008}; \citeproc{ref-Luque2016}{Luque
2017}) (e.g., individuals, alleles). This derivation partitions total
characteristic change into different components, making it possible to
isolate evolutionary mechanisms (e.g., selection) and levels of
biological organisation (e.g., group, individual)
(\citeproc{ref-Frank1995}{Frank 1995}, \citeproc{ref-Frank2012}{2012};
\citeproc{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
\citeproc{ref-Luque2016}{Luque 2017}; \citeproc{ref-Okasha2020}{Okasha
and Otsuka 2020}). Because of its abstract nature and lack of any
system-specific assumptions, the Price equation is not dynamically
sufficient and makes no predictions about what will happen in any
particular system (\citeproc{ref-Gardner2020}{Gardner 2020}). Its role
is not to predict, but to formally and completely define and separate
components of evolutionary change. The same is true of the general
equation for population change (eqns 2 and 3), at least as we have used
it here where it serves to define what population growth means in
ecology. This equation formally and completely describes population
change in terms of births and deaths. In eqn 1, we therefore have a
fundamental equation from which we can derive complete ecological and
evolutionary change in any closed biological population, but one that is
revealed to be necessarily abstract and not dynamically sufficient. We
anticipate that this will be useful for eco-evolutionary theory in the
same way that the Price equation is useful for evolutionary theory:
facilitating specific model development and identifying new conceptual
insights, unresolved errors, and sources of model disagreements.

Our unification recovers the equivalence between the finite rate of
increase \(\lambda\) (Box 2) (\citeproc{ref-Gotelli2001}{Gotelli 2001})
and population mean evolutionary fitness \(\bar{w}\) (Box 1). The
population growth equation \(N_{t+1} = N_{t}\lambda\) can therefore be
rewritten as \(N_{t+1} = N_{t}\bar{w}\). This specific equivalence has
been proposed before (\citeproc{ref-Lande1976}{Lande 1976}), as has the
broader relationship between population growth rate and evolutionary
fitness (\citeproc{ref-Fisher1930}{Fisher 1930};
\citeproc{ref-Charlesworth1980}{Charlesworth 1980};
\citeproc{ref-Lande1982}{Lande 1982}; \citeproc{ref-Roff2008}{Roff
2008}; \citeproc{ref-Lion2018}{Lion 2018}). We show this from first
principles and clarify the relationship between fitness and population
growth. Over an arbitrary length of time, fitness is properly defined as
\(w_{i} = \beta_{i} - \delta_{i} + 1\). Over an individual's lifetime
(which, by definition, includes death), fitness is therefore
\(\beta_{i}\). The rate of change in ecology and evolution are reflected
in the first and second statistical moments of fitness, respectively.
Population growth rate reflects mean fitness \(\bar{w}\), while the rate
of evolutionary change reflects the variance in fitness
\(\mathrm{Var}(w)/\bar{w}\) (i.e., Fisher's fundamental theorem)
(\citeproc{ref-Frank1997}{Frank 1997}; \citeproc{ref-Rice2004}{Rice
2004}; \citeproc{ref-Queller2017}{Queller 2017}).

Our unification may also help explain, at least partially, some of the
success of classical population genetic models. For decades, population
genetics (and to some extent quantitative genetics) has been accused of
being a reductionist view of evolution, reducing everything to changes
in allele frequencies and abstracting away from individuals and their
environments (the ecological interactions)
(\citeproc{ref-MacColl2011}{MacColl 2011}). This has been a line of
argumentation by some defenders of the so-called Extended Synthesis
(\citeproc{ref-Pigliucci2009}{Pigliucci 2009}), especially in relation
to niche construction (\citeproc{ref-Odling-smee2003}{Odling-Smee et al.
2003}). Famously, Mayr (\citeproc{ref-Mayr1960}{1959}) characterized
population genetics as a simple input and output of genes, analogous to
``the adding of certain beans to a beanbag and the withdrawing of
others'' (also called ``beanbag genetics''). Historical critics of
population genetics could not articulate a clear explanation for why it
works so well despite all of its idealisations and simplifications. From
the Price equation, we are able to recover classical population and
quantitative genetic models (\citeproc{ref-Queller2017}{Queller 2017})
and develop new ones (\citeproc{ref-Rice2004}{Rice 2004},
\citeproc{ref-Rice2020}{2020}; \citeproc{ref-Luque2016}{Luque 2017};
\citeproc{ref-Lion2018}{Lion 2018}). Our eqn 1 contains ecology at its
core, and we show how the Price equation logically follows from it after
accounting for absolute population growth (eqn 8). We therefore conclude
that population and quantitative genetic equations contain ecology (no
matter how hidden), and the ecological nature of evolution is implicit
in population and quantitative genetic models.

@@

Our fundamental equation is complete and exact. It therefore implicitly
includes any effects of density dependence on population growth (see Box
2), or any social effects on evolutionary change (see Box 1). Both of
these effects can be made explicit by specifying how other individuals
in a population affect birth and death of a focal individual. Assuming
that effects of individuals are additive, independent, and identical, we
can derive well-established logistic growth equations and Hamilton's
rule from eqn 1 (Supporting Information S1).

We have shown that we can derive the fundamental equations of population
ecology and biological evolution from a single unifying equation.
Lastly, we propose our eqn 1 as a potential starting point for defining
ecosystem function and further conceptual unification between ecology,
evolution, and ecosystem function. The Price equation has previously
been used to investigate ecosystem function
(\citeproc{ref-Loreau2001}{Loreau and Hector 2001};
\citeproc{ref-Fox2006}{Fox 2006}), but not with any attempt towards
conceptual unification with evolutionary biology. For example, Fox
(\citeproc{ref-Fox2006}{2006}) applied the abstract properties of the
Price equation to partition total change in ecosystem function into
separate components attributable to species richness, species
composition, and context dependent effects. This approach provides a
framework for comparing the effects of biodiversity on ecosystem
function in empirical systems (\citeproc{ref-Fox2006}{Fox 2006};
\citeproc{ref-Winfree2015}{Winfree et al. 2015};
\citeproc{ref-Mateo-Tomas2017}{Mateo-Tomás et al. 2017}). Instead, our
eqn 1 defines \(\Omega\) as total ecosystem function contributed by a
focal population. It is therefore possible to investigate ecological,
evolutionary, and ecosystem function change from the same shared
framework.

@@

\bibitem[\citeproctext]{ref-Baravalle2024}
Baravalle, L., A. Roffé, V. J. Luque, and S. Ginnobili. 2025.
\href{https://doi.org/10.1007/s13752-024-00482-4}{The value of price}.
Pages 12--24 \emph{in}Biological Theory (Vol. 20).

@@

@@
